High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-09-13 manuscript/Manuscript.tex

\authors{Yiming Zhang\affil{1}, Nicholas L. Swanson-Hysell\affil{1}, James D. Miller Jr.\affil{2}, Margaret S. Avery\affil{1,3}}

@@

\begin{keypoints}
\item New ca. 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field during the late Mesoproterozoic is inconsistent with previous interpretation of long term decay
\item A strong geodynamo like today likely persisted for about 25 myr during the Midcontinent Rift development
\end{keypoints}

@@

\begin{abstract}
New high-quality absolute paleointensity data from anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift (MCR) record a virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$ ca. 1092 Ma. This cooling rate corrected paleointensity estimate from an MCR intrusive unit is similar to previous results from extrusive units that span from ca 1108 Ma to ca. 1085 Ma. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a geomagnetic field strength persisted through the Midcontinent Rift development and the data are inconsistent with the interpretation of a decay of Earth's geodynamo strength through the late Mesoproterozoic.

@@

That we have a solid inner core today is the result of secular cooling and differentiation of the layered Earth. The timing of inner core formation during Earth's thermal evolution is amongst the major unsolved questions in Earth's history. Due to the uncertain constraints on the thermal conductivity of the core \cite{Konopkova2016a, Gomi2013a, Ohta2016a}, numerical simulations have given drastically different predictions on the inner core nucleation age, ranging from about half a billion years ago to prior to the Archean \cite{Pozzo2012a, Gubbins2004a, Nimmo2015a}. Modeling has suggested that the emergence of the solid inner core could have expressed as changes of Earth's surface magnetic field strength due to the onset of compositional convection between the solid inner core and the liquid outer core \cite<e.g.>{Aubert2009a}. Thus, absolute paleointensity records preserved by magnetic minerals in igneous rocks may provide us with a probe into the birth of the inner core. Acquiring these observational data is crucial for advancing our understanding of Earth's thermal evolution and for testing existing models on the commencement of the inner core growth. However, due to the scarcity of well-preserved ancient rocks and low success rates typical of paleointensity experiments, only 9\% of the records in the current global paleointensity database (\href{http://earth.liv.ac.uk/pint/}{PINT, \cite{Biggin2010a}}) are from before the Cambrian and the existing records sparsely span over 3 billion years of time (Fig. \ref{fig:all_PINT_data}).

\citeA{Bono2019a} filtered the available Precambrian paleointensity database (PINT) based on a set of quality criteria. A polynomial fit through the resulting paleointensity compilation from the Archean and the Proterozoic was used to interpret that a decrease trend of the estimated geomagnetic dipole strength occurred before the inner core started to form. In addition, \citeA{Bono2019a} recovered very low paleointensity values from ca. 565 Ma single silicate crystals  that were interpreted to be consistent with it reflecting the weak geodynamo after progressive cooling of the core before the onset of the rigorous compositional convection driven by the nucleation of the solid inner core. Such decay trend is recently supported by new low paleointensity estimates derived from the ca. 720 Ma Franklin Large Igneous Province \cite{Lloyd2021a}.

However, previous paleointensity data from the MCR rock units \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VADM values distinctly higher than what is predicted by the long term decay curve. These rock units notably are extrusive volcanics and span a tme period of about 25 myr during the development of the Midcontinent Rift magmatism. Many previous results from the Midcontinent Rift show non-ideal double-slope paleointensity behavior, which are difficult to interpret as they can yield two paleointensity estimates: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Nevertheless, single-slope, ideal behavior from the Lake Shore Traps \cite{Kulakov2013a} lava flows passed the selection criteria of \citeA{Bono2019a}, and a few lava flows yielded single-slope behavior from the Osler Volcanics \cite{Sprain2018a}. These data are inconsistent with the interpretation of a low geomagnetic field strength during the Midcontinent Rift because they yield significantly higher VADM estimates than that projected by the trend from \citeA{Bono2019a}.

To further enrich the Precambrian paleointensity database, and to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift, in this study we perform ``IZZI" paleointensity experiments \cite{Tauxe2004a} on a suite of anorthosite xenoliths hosted by the intrusive Beaver River diabase of the Beaver Bay Complex within the Midcontinent Rift.

@@

\section*{Geologic Settings}
The North American Midcontinent Rift (MCR) is a failed intracontinental rift where protracted magmatic activity lasted from \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}. Midcontinent Rift rocks extensively outcrop in today's Lake Superior region, with the total extent traceable by arcuate magnetic and gravity anomalies that extend to the southwest to Kansas, and to the southeast, to southern Michigan \cite{Hinze2020a}. Previous studies have divided magmatic activity in the rift into four stages based on interpreted changes in relative magmatic volume and the nature of magmatism: early ($\sim$1109–1104 Ma), latent ($\sim$1104–1098 Ma), main ($\sim$1098–1090 Ma) and late ($\sim$1090–1083 Ma) \cite{Vervoort2007a, Heaman2007a, Miller2013a}. In northeastern Minnesota, the Early Gabbro Series and the Felsic Series rocks of the Duluth Complex and reversed-polarity lavas of the lower North Shore Volcanic Group were emplaced during the early stage. The more voluminous Duluth Complex Layered Series and the plagioclase-rich Anorthositic Series, together with an associated $\sim$8 km thick extrusive volcanic sequences of the North Shore Volcanic Group (NSVG), were rapidly emplaced about 10 myr later at \textit{ca.} 1096 Ma during the main stage \cite{Paces1993a, Swanson-Hysell2020a}.

@@

\section*{Method}

\section*{Results}

\section*{Discussion}

\section*{Conclusion}

\acknowledgments
I thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help of instrumental operations, data interpretations, and research guidance. I thank the IRM for the generous U.S. Visiting Student Fellowship. I thank Nicholas Swanson-Hysell at UC Berkeley for advising this research. Margaret Avery provided field assistance.

>>>>

\authors{Yiming Zhang\affil{1}, Nicholas L. Swanson-Hysell\affil{1}, Margaret S. Avery\affil{1,3}}

@@

\begin{keypoints}
\item New ca. 1092 Ma Midcontinent Rift anorthosite record high paleointensity values
\item High VADM of the geomagnetic field is inconsistent with previous interpretation of Proterozoic long term decay
\item A strong geodynamo like today likely persisted for about 25 myr during the Midcontinent Rift development
\end{keypoints}

@@

\begin{abstract}
New high-quality absolute paleointensity data from anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift (MCR) record an average virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$ \textit{ca.} 1092 Ma. This cooling rate corrected paleointensity estimate from the MCR intrusive unit is similar to previous results from extrusive units that are dated to be \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosites can have abundant single-domain magnetic carriers that are ideal for paleointensity experiment, leading to a high success rate passing our selection criteria. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a geomagnetic field strength persisted through the Midcontinent Rift development and the data are inconsistent with the interpretation of a decay of Earth's geodynamo strength through the late Mesoproterozoic.

@@

That we have a solid inner core today is the result of secular cooling and differentiation of the layered Earth. The timing of inner core formation during Earth's thermal evolution is amongst the major unsolved questions in Earth's history. Due to the uncertain constraints on the thermal conductivity of the core \cite{Konopkova2016a, Gomi2013a, Ohta2016a}, numerical simulations have given drastically different predictions on the inner core nucleation age, ranging from about half a billion years ago to prior to the Archean \cite{Pozzo2012a, Gubbins2004a, Nimmo2015a}. Some models have suggested the thermal convection dominated liquid core motion experienced a long-term waning before the onset of the nucleation of the inner core, which kick-started the rigorous compositional convection between the solid inner core and liquid outer core \cite<e.g>{Davies2021a}. A corollary of such a prediction is that the geomagnetic axial dipole field, which is generated by the fluid motion in the liquid core, would have been decaying before the emergence of the solid inner core, when a significant increase in Earth's axial dipole field strength occurred \cite{Aubert2009a, Davies2021a}.

Paleomagnetic data from ancient rocks are one of the few types of observational data that bear on the long-term evolution of Earth’s core and can be used to test the predictions from numerical simulations. Paleointensity studies have found very low axial dipole moment (VDM) estimates from rocks formed in the late Ediacaran \cite{Bono2019a, Shcherbakova2019a, Thallner2021b}, followed by a recovery of strength near the  Ediacaran-Cambrian boundary. Other studies focusing on paleomagnetic directions have also found that the late Ediacaran might have been a period of time when the geomagnetic field were experiencing frequent reversals and waning of dipole dominance. It has been interpreted that these findings are consistent with the results of previous geodynamo models using high core conductivity values that suggest the inner core nucleation age being near the Ediacaran-Cambrian boundary. \citeA{Bono2019a} filtered previous data compiled within the PINT database based on a set of quality criteria. A polynomial fit through the resulting paleointensity compilation from the Archean and the Proterozoic was used to interpret that a decrease trend of the estimated geomagnetic dipole strength persisted through the Proterozoic. In addition, \citeA{Bono2019a} recovered very low paleointensity values from ca. 565 Ma single silicate crystals that were interpreted to be consistent with it reflecting the weak geodynamo near the Ediacaran-Cambrian boundary due to progressive cooling of the core before the onset of the rigorous compositional convection driven by the nucleation of the solid inner core. Such decay trend in the Neoproterozoic is recently supported by new low paleointensity estimates derived from the ca. 720 Ma Franklin Large Igneous Province \cite{Lloyd2021a}.

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today and are inconsistent with the prediction of the long term decay curve (Fig. \ref{}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using the ca. 1108 Ma Osler Volcanics, Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, and the Baraga and Marquette dikes, as well as the ca. 1096 Ma North Shore Volcanic Group. \citeA{Sprain2018a} obtained more high-quality paleointensity data using the ca. 1108 Ma Osler Volcanics and the ca. 1100 Ma Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years. In addition, an average VADM of 59 $\pm$ 11 ZAm$^2$ obtained \citeA{Kulakov2013a} using the ca. 1085 Ma Lake Shore Traps volcanics suggests that the present-day-like geomagnetic field strength likely persisted through the late stage of Midcontinent Rift magmatism. Taken together, these rock units that have been dated to span $\sim$25 myr during the development of the Midcontinent Rift magmatism \cite{Swanson-Hysell2019a} may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

But the non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength whereas \citeA{Kulakov2013a} used the high-temperature slop. Such non-ideal behavior were interpreted as failure by \siteA{Sprain2018a} who applied stricter paleointensity selection criteria and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. Similar concern on paleointensity result quality led \citeA{Bono2019a} to exclude all paleointensity records from the Midcontinent Rift but those from \citeA{Kulakov2013a}. Nevertheless, specimens that yield dominantly single-slope paleointensity results from the ca. 1105 Ma Osler Volcanic Group, Mamainse Point volcanics \cite{Sprain2018a}, as well as the ca. 1085 Ma Lake Shore Traps \cite{Kulakov2013a} suggest VADM values similar to that of the present-day values \cite{Sprain2018a}, inconsistent with the interpretation of a low geomagnetic field strength during the Midcontinent Rift (Fig. \ref{}).

Although it is typical of paleointensity experiments to suffer from low success rates, quenched volcanic glass \cite{Tauxe2004a} and single silicate minerals \cite{Tarduno2005a} are found to be able to produce high success rates, likely due to the magnetic carriers being dominated by single-domain magnetite and being well shielded by their host against thermochemical alteration during paleointensity experiments. In this study, we present new paleointensity results from ``IZZI" absolute paleointensity experiments \cite{Yu2004a} on a suite of ca. 1092 Ma anorthosite xenoliths and their diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. The anorthosite lithology yielded exceptionally high success rate thanks to the plagioclase crystals shielding the magnetic minerals within. With more single-slope, high quality paleointensity data we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 myr of Midcontinent Rift magmatic activity.

@@

\section*{Geologic Settings}
% a map of the lake superior including the pesonen units, the lake shore traps, the BRD, and the osler and mamainse, together with date compilation.

The North American Midcontinent Rift (MCR) is a failed intracontinental rift where protracted magmatic activity lasted from \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}. Previous studies have divided magmatic activity in the rift into four stages based on interpreted changes in relative magmatic volume and the nature of magmatism: early ($\sim$1109–1104 Ma), latent ($\sim$1104–1098 Ma), main ($\sim$1098–1090 Ma) and late ($\sim$1090–1084 Ma) \cite{Vervoort2007a, Heaman2007a, Miller2013a}. In particular, the main stage magmatism is characterized by rapid and punctuated events which resulted in voluminous emplacement of intrusives as well as coeval outpouring of extrusives, as can be exampled by the massive ca. 1096 Ma Duluth Complex and the North Shore Volcanic Group \cite{Swanson-Hysell2020a}, the ca. 1092 Ma Beaver Bay Complex and the Portage Lake Volcanics (Fig. \ref{}; \citeA{Zhang2021a}. Excellent rock preservation thanks to minimal post-rift alteration in the Midcontinent Rift have yielded high-quality paleomagnetic directional data which have now been paired with high-precision zircon geochronology to reconstruct the well-resolved apparent polar wander path for Laurentia (cratonic North America) called the Keweenawan Track (Fig. \ref{}; \citeA{Swanson-Hysell2019a}).

@@

\section*{Method}
\subsection*{Sample collection}
Sample cores were collected using a hand-held drill and were ori- ented using a magnetic compass as well as a sun compass when possible. Palaeohorizontal was determined from intercalated sedi- mentary layers and flow tops. 6–10 cores were drilled from each lava flow in order to robustly determine an accurate site mean. More detailed information about the sites and their context along with directional results are reported in Swanson-Hysell et al. (2009, 2014a,b). For palaeointensity analysis, 5–8 samples per flow were chosen. Samples were chosen based on demagnetization behaviour with a preference for samples wherein relatively little remanence is held by haematite such that their magnetization is dominated by (titano)magnetite recording a primary thermal remanent mag- netization. Haematite can be a significant carrier of remanence in Mamainse Point basalts (see Swanson-Hysell et al. 2011) and other successions around the Midcontinent Rift. Given that haematite can form at the expense of magmatic (titano)magnetite, such flows were deemed to not be appropriate targets for the palaeointensity experiments.

\subsection*{Paleointensity experiment}
A total of 133 specimens from 21 selected sites underwent palaeoin- tensity experiments that followed the stepwise double-heating Thel- lier method (Thellier & Thellier 1959), using the IZZI protocol (Tauxe & Staudigel 2004). pTRM checks were performed systemati- cally throughout the experiment to test whether there was significant mineralogical alteration due to heating and were assessed using the SCAT parameter of Shaar & Tauxe (2013). All remanence measure- ments were made on a 2G Enterprises DC-SQUID superconduct- ing rock magnetometer equipped with an automated pick-and-place sample changer system at the UC Berkeley Paleomagnetism labora- tory. The magnetometer is housed inside a three-layer magnetostatic shield that maintains background fields of less than 500 nT. Heating steps were performed using an ASC TD-48SC thermal demagne- tizer with a controlled field coil that allows for a magnetic field to be generated in the oven in conjunction with a DC power sup-ply. The thermal demagnetizer was degaussed with an alternating field following ‘in-field’ steps such that residual fields were <10 nT during ‘zero-field’ steps. Samples were placed in the same location within the thermal demagnetizer for each heating step and were maintained in the same orientation with regard to the applied field. During each heating step, samples remained at peak temperatures for 20 min. An applied laboratory field of 30 μT was used for all in- field steps. All heating steps were performed in air. The temperature increments for the experiments were chosen to cover characteristic remanent magnetizations held by (titano)magnetite, with smaller increment temperature steps performed close to the expected un- blocking temperature of stoichiometric magnetite. Hysteresis mea- surements were conducted at the Institute for Rock Magnetism at the University of Minnesota. Major hysteresis loops were mea- sured at room temperature using a Micromag Princeton Measure- ments vibrating sample magnetometer with nominal sensitivity of 5 × 10−9 Am2.

A total of 59 specimens from 13 anorthosite sites and 46 specimens from 7 diabase sites underwent paleointensity experiments that followed the stepwise double-heating Thellier method with temperature steps up to 585 $^\circ$C \cite{Thellier1959a, Tauxe2004a}. On top of the IZZI modification of the Thellier experimental protocol, we also performed a comparative study where we added an extra step of 20 mT alternating field (AF) cleaning on half of the specimens after each in-field step. The purpose is to study whether the AF step could help improve specimens' behavior during paleointensity experiments. pTRM checks were performed systematically throughout the experiment to test whether there was significant mineralogical alteration due to heating and were assessed using the SCAT parameter of \citeA{Shaar2013a}. All remanence measurements were made on a 2G Enterprises DC-SQUID superconducting rock magnetometer equipped with an automated sample changer system in the paleomagnetism lab at UC Berkeley. Heating steps were performed using an ASC TD-48SC thermal demagnetizer with a controlled field coil that allows for a magnetic field to be generated in the oven in conjunction with a DC power supply. The thermal demagnetizer was degaussed with an alternating field following in-field steps such that residual fields were $<$10 nT during zero-field steps. Samples were placed in the same location within the thermal demagnetizer for each heating step and were maintained in the same orientation with regard to the applied field. During each heating step, samples remained at peak temperatures for 20 min. An applied laboratory field of 30 $\mu$T was used for all in-field steps. All heating steps were performed in air. The temperature increments for the experiments were chosen to cover characteristic remanent magnetizations held by (titano)magnetite, with smaller increment temperature steps performed close to the expected unblocking temperature of stoichiometric magnetite.

\subsection*{Paleointensity result selection}
The following criteria were used as quality filters on the palaeoin- tensity results: (1) a maximum angular deviation (MAD; Kirschvink 1980) of <20◦; (2) scatter parameter (β; Coe et al. 1978) values of <15 per cent; (3) a deviation angle (DANG; Tanaka & Kobayashi 2003; Tauxe & Staudigel 2004) of < 5◦; (4) fraction of remanence (FRAC; Shaar & Tauxe 2013) >0.6; (5) scatter statistic (SCAT; Shaar & Tauxe 2013) = TRUE; (6) a maximum gap (GAP-Max; Shaar & Tauxe 2013) < 0.6; (7) number of pTRM checks > 2; (7) and number of measurements used for palaeointensity determi- nation ≥ 4; (Table 1)

The following criteria were used as quality filters on the paleointensity results: (1) a maximum angular deviation (MAD; \citeA{Kirschvink1980a}) of $<$ 20; (2) scatter parameter ($\beta$; citeA{Coe1978a}) values of $<$ 15 \%; (3) a deviation angle (DANG; \citeA{Tanaka2003a}; \citeA{Tauxe2004a} of $<$ 5; (4) fraction of remanence (FRAC; \citeA{Shaar2013a}) $>$ 0.6; (5) scatter statistic (SCAT; \citeA{Shaar2013a}) = TRUE; (6) a maximum gap (GAP-Max; \citeA{Shaar2013a}) $<$ 0.6; (7) number of pTRM checks $>$ 2; (7) and number of measurements used for paleointensity determination $\geq$ 4; (Table \ref{tab:criteria}). The MAD measures the scatter about the best-fit line through NRM steps in the selected interval for which the intensity is defined. DANG, the deviation angle, is the angle between the best-fit direction that is free floating and the direction between the centre of mass of the data and the origin of the vector component diagram (\citeA{Tanaka2003a}; \citeA{Tauxe2004a}). Both MAD and DANG assess the directional variation of the NRM, with MAD measuring the scatter in the NRM directions and DANG assessing whether the component is trending toward the origin of the Zijderveld plot. $\beta$ is the ``scatter" parameter of \citeA{Coe1978a} and is the ratio of the standard error of the slope of the best-fit line of the selected NRM and pTRM points on an Arai plot to the absolute value of the slope. FRAC is the fraction of the NRM that is used in the best-fit line \cite{Shaar2013a}. The FRAC value was chosen to preferentially select samples with dominantly single-slope Arai plots. GAP-Max is the maximum gap between two points on the Arai plot determined by vector arithmetic. SCAT is a Boolean operator which uses the error on the best-fit slope of the selected data on the Arai plot to determine if the data are overly scattered. The parameter is used to assess pTRM checks in addition to assessing the degree to which IZZI steps are zigzagged. $\beta$, FRAC, GAP-Max and SCAT are all statistics to assess the behavior of Arai plots. See the Standard Paleointensity Definitions (\citeA{Paterson2014a}, https://earthref.org/PmagPy/SPD/home.html, last accessed 2018 March 17) for more details. Data analysis was conducted using Thellier GUI \cite{Shaar2013a} within the PmagPy software package \cite{Tauxe2016a}.

\subsection*{Rock magnetic experiment}

Hysteresis, backfield, and magnetic property measurements were conducted at the Institute for Rock Magnetism (IRM) at the University of Minnesota. Major hysteresis loops and backfield curves were measured at room temperature using a Micromag Princeton Measurements vibrating sample magnetometer with nominal sensitivity of $5 \times 10^{-9}$ Am\textsuperscript{2}. We used the magnetic property measurement system (MPMS) at the IRM to aid in the identification of magnetic mineral. The operational sequence follows an in-field cooling step in a 2.5 T field from room temperature to 10 K, a zero-field cooling from room temperature to 10 K, and a final cooling from room temperature to 10 K after an IRM acquisition at room temperature.

Backfield demagnetization experiments were conducted and used to develop coercivity spectra (Fig. \ref{fig:backfield}). The spectra were subsequently modeled to fit for the distributions of different  populations of magnetic particles using similar procedure as \cite{Maxbauer2016a}. Most spectra can be well approximated with models with one component or two components with overlapping coercivity ranges (Fig. \ref{fig:backfield}). The dominance of single-component distributions from the unmixed coercivity spectra likely suggests minimal alterations or formation of secondary magnetic mineral within the samples. By compiling the values of median destructive field (MDF) from all measured specimens and categorizing them in terms of their sister specimens' paleointensity results, I found that the anorthosites samples that pass paleointensity selection criteria have distinctively higher MDF values than those did not (Fig. \ref{fig:backfield}), and all diabase specimens have low MDF values like the non-ideal anorthosites (Fig. \ref{fig:backfield}). The higher MDFs ($\sim$60 mT) of the anorthosites are similar to what is typical of stoichiometric, single-domain magnetite grains which are favored for paleointensity experiments. Furthermore, the low MDFs displayed in other specimens could be linked to a dominant population of large interstitial Fe-Ti oxides that are more likely to form multi-domain grains, which have been observed in both the anorthosite and diabase (Fig. \ref{fig:field}). In addition, preliminary tests on single anorthosite crystals might suggest that not all single crystals are necessarily better for paleointensity research when comparing to bulk samples, given their similar low MDF to the failed diabase and anorthosite specimens. However, more single crystal grains with paired paleointensity data and rock magnetic data may be needed before a general conclusion can be drawn.

\section*{Results and Discussion}
\subsection*{Paleomagnetism}
The paleodirectional results of AF demagnetization of 7 diabase sites and thermal demagnetization of 16 anorthosite sites are shown in Table \ref{tab:}. In some cases, specimens from both lithologies show a minimal remanence component that was removed during very low AF field ($<$ 10 mT) for diabase and during low temperature demagnetization steps for anorthosites ($<$ 300 $^\circ$C). This component is likely a modern day field overprint, since its direction resembles the present-day field direction. The dominant components in both lithologies are northwest and down (Figure \ref{fig:directions}). This is likely the primary remanent magnetization recorded by the diabase and anorthosite during the emplacement of the diabase, since it is origin-trending and the principal component fitting yielded good specimen-level and site-level consistence. Common mean test (\cite{McFadden1990a}) results based on site-level directional comparison between anorthosites and their in situ diabase hosts are shown in Figure \ref{fig:directions}. Most anorthosite-diabase pairs have tightly clustered directions, and the three anorthosite sites that do no pass the statistical test with their diabase hosts nevertheless have within-site coherent directions that are close to the diabase directions (Figure \ref{fig:directions}). Putting the grand mean paleodirections and inferred virtual geomagnetic poles (VGPs) of both lithologies together in context of a previously synthesized apparent polar wander path (APWP) from the volcanics of the Midcontinent Rift \cite{Swanson-Hysell2019a}, we find the pole positions recovered from both the diabase and anorthosite to be lying close to the estimated VGP position around 1095 Ma, which is consistent with the geochronology constraints (between 1093.94 $\pm$ 0.28 Ma and 1091.61$\pm$ 0.14 Ma, \cite{Swanson-Hysell2019a}).

Paleointensity experiment results show three types of dominant specimen behaviors on Arai plots (Figure \ref{fig:Arai_plot}): dominantly single-slope (--- out of --- specimens), double- slope (--- out of --- specimens) and zigzagging (--- out of --- specimens) (Figure \ref{fig:Arai_plot}). It is worth mentioning that almost all specimens have single direction demagnetization result as in Zijderveld plots of all zero-field steps. Upon applying the paleointensity result selection criteria (Table \ref{tab:}), we found a total of 31 out 59 anorthosite specimens and 4 out of 46 diabase specimens were accepted, yielding 7 anorthosite sites and 1 diabase sites with interpretable site-level paleointensity results (Table \ref{}). The selection criteria successfully rejected double-slope results (FRAC $>$ 0.6) and zigzagging results ($\beta$, DANG, and SCAT parameter). In addition, some rejected specimens have poor pTRM checks especially at high temperature steps (Figure \ref{fig:Arai_plot}). This may suggest a high degree of alteration in magnetic mineralogy during heating steps. The accepted anorthosite specimens do display straight Arai plots and the accepted fractions of temperature steps are mostly covering the primary remanence components (Figure \ref{fig:Arai_plot}). In terms of the diabase specimens, all but four of them failed the selection because double-sloping Arai plots, poor pTRM checks, and zigzagging behaviors (Figure \ref{fig:Arai_plot}). Even specimens in site BD2, the only site that passed selection, do not always display ideal Arai plots (Figure \ref{fig:Arai_plot}). Additionally, all accepted results from site BD2 are from low-temperature fractions of Arai plots (below 500 $^\circ$C), which are different than those used in anorthosite sites (typically above 500 $^\circ$). Although the lower temperature fraction of the Arai plots are straight and pTRM checks are consistent (Figure \ref{fig:Arai_plot}), we have to acknowledge the existence of second slopes at high temperature fractions (Figure \ref{fig:Arai_plot}). Nevertheless, the interpreted ancient geomagnetic field value from the accepted slopes is very similar to those reported by anorthosites (Figure \ref{fig:Arai_plot}).

In this study, it is particularly important to assess the cooling rate effects, remanence anisotropy, and potential non-linear TRM or pTRM acquisition characteristics of diabase and anorthosites. Because the cooling duration of intrusive rocks could be much longer than the extrusive rocks, the slowly cooled diabase and anorthosite can bias laboratory paleointensity estimates toward an overestimate for SD grains \cite{Dodson1980a, Halgedahl1980a, Selkin2000a} or underestimation for MD grains \cite{McClelland-Brown1984a}. Further, preferred orientations of ferromagnetic fabrics along plagioclase crystallographic orientations in anorthosites could introduce a large degree of anisotropy in remanent magnetizations \cite{Selkin2007a, Biedermann2018a}. \citeA{Selkin2007a} reported that large, acicular SD grains can have profound non-linear acquisitions of TRM. According to their experiments, grains with such characteristics that are hosted within silicate crystals can saturate in fields as low as earth's field \cite{Selkin2007a}.

Backfield experiment results suggest that the dominant remanence carrier in accepted anorthosite sites are likely SD magnetite (Figure \ref{fig:MDF}). The median destructive fields (MDF) of all accepted anorthosites are distinctively higher than those of the rejected anorthosites and the diabase. Moreover, the median of the MDFs of accepted anorthosites is about 50 mT, which is in the expected range of typical no-interactive SD magnetite grains. In addition, the anorthosites' narrow unblocking temperatures near the magnetite curie temperature also support that we idealize the remanence carriers in anorthosites to be SD magnetite.

From the thermal history model we can also estimate the time duration which the diabase and anorthosite spent to block in the primary component of their remanence. We found the cooling time to be on the order of magnitude of $\sim$1 kyr from 580 to 500 $^\circ$C, which corresponds to a cooling rate of $1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at about $1.3\times10^{-1}$ $^\circ$C s$^{-1}$ if we assume it takes 10 minutes in lab condition to cool through the same temperature interval. As a result, this significant cooling rate difference leads to a prediction of 35.5\% overestimate of true field \cite{Halgedahl1980a}. We therefore correct our paleointensity results by a factor of 0.74.

It has been well illustrated through experiments on anorthosites that anisotropic remanence acquisition can have a significant impact on paleointensity estimates \cite{Selkin2000a}. Following similar experiment protocol, we found that at least the bulk anorthosite specimens in this study are not very anisotropic. This result can be further supported by the fact that the recovered paleodirectional data from the anorthosite closely match those from the paired diabase hosts (Figure \ref{fig:directions}), the lithology of which is usually considered isotropic in terms of recording remanence magnetization. Both the diabase and anorthosite display consistent interpreted VGPs at the time of emplacement that matche with the expected position based on the volcanic APWP record (Figure \ref{fig:poles}, \citeA{Swanson-Hysell2019a}). We therefore do not correct our samples for anisotropy of remanence.

We also performed non-linear full TRM checks following the experiment protocol proposed by \citeA{Selkin2007a}. We imparted full TRMs on our anorthosite specimens (after the paleointensity experiment) by heating them up to 600 $^\circ$C and cooling in field of 30, 50, 70, and 90 $\mu$T respectively. Unlike the anorthosites studied by \citeA{Selkin2007a}, none of our specimens show a saturating trend at progressively higher fields. More, the normalized magnetizations show nearly linear relationship with lab field strengths. There is thus no need for correcting non-linear thermal remanence acquisitions.

Figure \ref{fig:PINT_plot} shows all accepted anorthosite and diabase specimen paleointensity results and their site-level means with one standard deviation envelopes. An anorthosite site mean paleointensity weighted by site standard deviations is plotted as a dashed orange line, and the standard deviation envelope as dashed green lines. The anorthosites are also grouped by their geographical vicinities \ref{fig:PINT_plot}. The site-level results are shown in Table \ref{tab:PINT_result}. Anorthosite sites AX10, 11, 12, 13, 16 yielded very high within-site coherence (Figure \ref{fig:Arai_plots, fig:PINT_plot}), which suggest high-confidence site-level paleointensity estimates. Clearly there are two populations of paleointensity results: a group of high estimates from anorthosite site 11 and 13 which have very high within-site consistency and are in agreement with the results from BRD site 2; and another group of low estimates from anorthosite site 6, 8, 10, 12, and 16, which also have tight agreement across sites.

\subsection*{Rock magnetism}
\begin{figure}[h!]
\noindent\includegraphics[width=0.5\textwidth]{manuscript/Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{ }}
\label{fig:PINT_cooling_corrected}
\end{figure}

% that this study has robust data, discuss the reasoning using rock mag data that reflect the success and failure in PINT experiment

% cooling rate correction problem - that the Nagy et al paper can suggest even less of a correction and higher of the estimated field. theoretical thermal modeling by \citeA{Zhang2021a} suggest tha the anorthosite xenoliths were likely heated to diabase magma temperature regardless of their sizes, and cooled through the typical characteristic unblocking temperature range from 580 to 500 on the order of 1000 years. Comparing to the typical lab cooling time of about 100 seconds duirng paleointensity experiments, the significantly slower cooling rate when the NRM was acquired as the anorthosite xenoliths cooled with the diabase led to overestimate of paleointensity by up to 40\%. Overall range of correction factor can be 1.1-1.4, in any case consistent with a high field strength

% that this dataset also could have suffered from problem of averaging out secular variations, given that the mean pmag directions deviate from what is expected from the sythesized APWP, nevertheless the variations in estimate absolute paleointensity lead us to interpret that albeit close, the they intensity data were acquired at distinctly different times that are sufficiently far apart such that the surface geomagnetic field had already changed.

% that \citeA{bono2019a} proposed an interesting trend but has problems in the directions, likely to be a short term feature of a low field during a reversal, instead of a low field as a result of long term cooling.

% that Lloyd 2021 also presented low field that fits the trend, but they similarly have the issue of reversal, and that they do not have good control on secular variations

% Together with previously published data, the Midcontinent Rift rocks are recording 1.1 Ga earth geomagnetic field that is stronger than the mean of the past 300 Ma (Fig. \ref{}).

\section*{Conclusion}
high success rate comparing to volcanics (more than half success sites), producing good consistent site level results, and across sites report present-day like field like today.

AF pre-treatment is helpful in that it increase the specimens/sites passing the selecting criteria, suggesting that the method likely helped improve the MD ferromagnetic carriers.

Rock mag and petrographic data might suggest differences in terms of the value of peak field on coercivity spectra between those passed and those didn't.

IRM measured before and after an acid dunk in --- concentration acid on specimens after IZZI experiment show a difference (significant drop) maybe? between specs that passed and those did not.

geographical association

Anyways. anisotropy of full TRM experiment on accepted specs show low degree of bulk sample scale full TRM anisotropy, with a range of ---. Cooling rate correction based on \cite{Halgedahl1980a} suggest a lab over estimate of B ancient of 35 \%, assuming constant cooling rate in natural condition. The cooling rate corrected field is $\sim$40 $\mu$T which is similar to today's field strength at similar latitude.
